# Student's $t$-test {#ram-t}

In this section,
the Student's $t$-test is presented as a structural equation model
using the RAM notation. 
Let $y$ be a continuous dependent variable,
$x$ be a dichotomous independent variable
$\left( x = \{0, 1\} \right)$,
and $\varepsilon$ be the stochastic error term
with mean 0 and constant variance of $\sigma_{\varepsilon}^{2}$
across the values of $x$.
The associations of the variables are given by

\begin{equation*}
  y
  =
  \alpha + \beta x + \varepsilon
\end{equation*}

\noindent where

- $\alpha$ is the expected value of $y$ when $x = 0$
- $\beta$ is the unit change in $y$ for unit change in $x$
- $\alpha + \beta$ is the expected value of $y$ when $x = 1$

```{tikz, ram-t-01, echo = FALSE, fig.width = 4, fig.cap = "Student's $t$-test", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", cache = TRUE}
\usetikzlibrary{er, arrows, positioning}
\begin{tikzpicture}[
  auto,
  node distance = 15mm,
  manifest/.style = {
    rectangle,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  latent/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 15mm,
    align = center
  },
  inv/.style = {
    rectangle,
    draw=none,
    fill=none,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  error/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 7mm,
    align = center
  },
  mean/.style={
    regular polygon,
    regular polygon sides = 3,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  path/.style = {
    ->,
    thick,
    >=stealth'
  },
  cov/.style = {
    <->,
    thick,
    >=stealth'
  },
]
  \node[manifest] (y) {$y$};
  \node[manifest] (x) [right = of y] {$x$};
  \node[mean] (1) [above = of y] {1};
  \node[error] (varepsilon) [left = of y] {$\varepsilon$};
  \draw[path] (1) to node[right] {$\alpha$} (y);
  \draw[path] (1) to node[above] {$\mu_x$} (x);
  \draw[path] (x) to node[above] {$\beta$} (y);
  \draw[path] (varepsilon) to node[above] {$1$} (y);
  \draw[cov] (1) to[out=70,in=110,looseness=5] node[above] {$1$} (1);
  \draw[cov] (x) to[out=70,in=110,looseness=5] node[above] {$\sigma_{x}^{2}$} (x);
  \draw[cov] (varepsilon) to[out=70,in=110,looseness=5] node[above] {$\sigma_{\varepsilon}^{2}$} (varepsilon);
\end{tikzpicture}
```

## Symbolic

Let $\left\{ y, x, \varepsilon \right\}$ be the variables of interest.

```{r, echo = FALSE}
v <- matrix(
  data = c(
    paste(expression(alpha + beta * mu[x])),
    paste(expression(mu[x])),
    0
  )
)
u <- matrix(
  data = c(
    paste(expression(alpha)),
    paste(expression(mu[x])),
    0
  )
)
A <- S <- matrix(
  data = 0,
  nrow = 3,
  ncol = 3
)
I <- diag(dim(A)[1])
A[1, 2] <- paste(expression(beta))
A[1, 3] <- 1
diag(S) <- c(
  0,
  paste(expression(sigma[x]^2)),
  paste(expression(sigma[varepsilon]^2))
)
filter <- matrix(
  data = c(
    1,
    0,
    0,
    1,
    0,
    0
  ),
  nrow = 2,
  ncol = dim(A)[2]
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{A} &=",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{S} &=",
  Ryacas::tex(Ryacas::ysym(S)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{C} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{S} \\left[ \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\right]^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  "\\mathbf{E} \\mathbf{S} \\mathbf{E}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A))),
  Ryacas::tex(Ryacas::ysym(S)),
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A))),
  "^{\\mathsf{T}}",
  "\\\\ &=",
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A)) * Ryacas::ysym(S) * t(solve(Ryacas::ysym(I) - Ryacas::ysym(A)))),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{F} &=",
  Ryacas::tex(Ryacas::ysym(filter)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{M} &=",
  "\\mathbf{F} \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{S} \\left[ \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\right]^{\\mathsf{T}} \\mathbf{F}^{\\mathsf{T}}",
  "\\\\ &=",
  "\\mathbf{F} \\mathbf{E} \\mathbf{S} \\mathbf{E}^{\\mathsf{T}} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  "\\mathbf{F} \\mathbf{C} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(Ryacas::ysym(filter)),
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A)) * Ryacas::ysym(S) * t(solve(Ryacas::ysym(I) - Ryacas::ysym(A)))),
  Ryacas::tex(Ryacas::ysym(filter)),
  "^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(Ryacas::ysym(filter) * solve(Ryacas::ysym(I) - Ryacas::ysym(A)) * Ryacas::ysym(S) * t(solve(Ryacas::ysym(I) - Ryacas::ysym(A))) * t(Ryacas::ysym(filter))),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{v} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{u}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]^{\\mathsf{-1}}",
  Ryacas::tex(Ryacas::ysym(u)),
  "\\\\ &=",
  Ryacas::tex(ramR::v_sym(A, u)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{u} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right) \\mathbf{v}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]",
  Ryacas::tex(Ryacas::ysym(v)),
  "\\\\ &=",
  Ryacas::tex(ramR::u_sym(A, v)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{g} &=",
  "\\mathbf{F} \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{u}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]^{-1}",
  Ryacas::tex(Ryacas::ysym(u)),
  "\\\\ &=",
  Ryacas::tex(ramR::g_sym(A, u, filter)),
  "\\end{align*}",
  sep = ""
)
```

### Using the `ramR` Package

```{r}
A
```

```{r}
S
```

```{r}
u
```

```{r}
filter
```

The covariance expectations
can be symbolically derived using the `ramR::C_sym()` function.

```{r}
ramR::C_sym(A, S)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{C} =",
  Ryacas::tex(ramR::C_sym(A, S)),
  "\\end{equation*}",
  sep = ""
)
```

The covariance expectations for the observed variables
can be symbolically derived using the `ramR::M_sym()` function.

```{r}
ramR::M_sym(A, S, filter)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{M} =",
  Ryacas::tex(ramR::M_sym(A, S, filter)),
  "\\end{equation*}",
  sep = ""
)
```

The mean expectations
can be symbolically derived using the `ramR::v_sym()` function.

```{r}
ramR::v_sym(A, u)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{v} =",
  Ryacas::tex(ramR::v_sym(A, u)),
  "\\end{equation*}",
  sep = ""
)
```

The mean expectations for the observed variables
can be symbolically derived using the `ramR::g_sym()` function.

```{r}
ramR::g_sym(A, u, filter)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{g} =",
  Ryacas::tex(ramR::g_sym(A, u, filter)),
  "\\end{equation*}",
  sep = ""
)
```

## Numerical Example

```{r, echo = FALSE}
set.seed(42)
n <- 1000000
group_0 <- cbind(
  y = rnorm(n = n),
  constant = 1,
  x = 0
)
group_1 <- cbind(
  y = rnorm(n = n, mean = 1),
  constant = 1,
  x = 1
)
data <- rbind(group_0, group_1)
colnames(data) <- c(
  "y",
  "c",
  "x"
)
beta <- matrixR::LS(
  data[, 2:3],
  data[, 1]
)
data <- cbind(
  data[, -2],
  epsilon = as.vector(data[, 2:3] %*% beta)
)
df <- as.data.frame(data[, 1:2])
mu <- as.matrix(colMeans(data))
mu_x <- mu[2, 1]
Sigma <- var(data)
s <- diag(Sigma)
alpha <- beta[1, 1]
beta <- beta[2, 1]
a1 <- c(
  0,
  beta,
  1
)
A <- S <- matrix(
  data = 0,
  nrow = dim(mu)[1],
  ncol = dim(mu)[1]
)
A[1, ] <- a1
diag(S) <- s
S[1, 1] <- 0
sigma2x <- S[2 , 2]
sigma2e <- S[3 , 3]
v <- matrix(
  data = c(
    alpha + beta * mu_x,
    mu_x,
    0
  )
)
u <- matrix(
  data = c(
    alpha,
    mu_x,
    0
  )
)
rownames(u) <- rownames(v) <- rownames(A) <- rownames(S) <- colnames(A) <- colnames(S) <- colnames(filter) <- c(
  "y",
  "x",
  "e"
)
rownames(filter) <- c(
  "y",
  "x"
)
```

```{r}
head(df)
summary(df)
```

```{r, echo = FALSE, eval = FALSE}
var(df)
var(df[which(df$x == 0), ][1])
var(df[which(df$x == 1), ][1])
```

```{r, echo = FALSE}
boxplot(
  y ~ x,
  data = df
)
hist(
  df[which(df$x == 0), ][["y"]],
  xlab = "y",
  main = "Histogram of y for x = 0."
)
hist(
  df[which(df$x == 1), ][["y"]],
  xlab = "y",
  main = "Histogram of y for x = 1."
)
```

```{r}
t.test(y ~ x, data = df)
```

```{r}
summary(lm(y ~ x, data = df))
```

```{r}
model <- "
  y ~ x
  y ~ 1
  x ~ 1
"
fit <- lavaan::sem(model, data = df)
lavaan::summary(fit)
```

```{r, echo = FALSE}
parameter <- round(
  c(
  unname(alpha),
  unname(beta),
  unname(sigma2x),
  unname(sigma2e),
  unname(mu_x)
), 
digits = 2
)
label <- c(
  "$\\alpha$",
  "$\\beta$",
  "$\\sigma^{2}_{x}$",
  "$\\sigma^{2}_{\\varepsilon}$",
  "$\\mu_x$"
)
knitr::kable(
  cbind(
    label,
    parameter
  )
)
```

### Using the `ramR` Package

```{r}
A
```

```{r}
S
```

```{r}
u
```

```{r}
filter
```

The covariance expectations
can be numerically derived using the `ramR::C_num()` function.

```{r}
ramR::C_num(A, S)
```

The covariance expectations for the observed variables
can be numerically derived using the `ramR::M_num()` function.

```{r}
ramR::M_num(A, S, filter)
```

The mean expectations
can be numerically derived using the `ramR::v_num()` function.

```{r}
ramR::v_num(A, u)
```

The mean expectations for the observed variables
can be numerically derived using the `ramR::v_num()` function.

```{r}
ramR::g_num(A, u, filter)
```
