# Student's $t$-test {#ram-t}

In this section,
the Student's $t$-test is presented as a structural equation model
using the RAM notation.
Let $y$ be a continuous dependent variable,
$x$ be a dichotomous independent variable
$\left( x = \{0, 1\} \right)$,
and $\varepsilon$ be the stochastic error term
with mean 0 and constant variance of $\sigma_{\varepsilon}^{2}$
across the values of $x$.
The associations of the variables are given by

\begin{equation*}
  y
  =
  \alpha + \beta x + \varepsilon
\end{equation*}

\noindent where

- $\alpha$ is the expected value of $y$ when $x = 0$
- $\beta$ is the unit change in $y$ for unit change in $x$
- $\alpha + \beta$ is the expected value of $y$ when $x = 1$

```{tikz, ram-t-01, echo=FALSE, fig.width=4, fig.cap="Student's $t$-test", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", cache=TRUE}
\usetikzlibrary{er, arrows, positioning}
\begin{tikzpicture}[
  auto,
  node distance = 15mm,
  manifest/.style = {
    rectangle,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  latent/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 15mm,
    align = center
  },
  inv/.style = {
    rectangle,
    draw=none,
    fill=none,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  error/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 7mm,
    align = center
  },
  mean/.style={
    regular polygon,
    regular polygon sides = 3,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 7mm,
    minimum height = 7mm
  },
  path/.style = {
    ->,
    thick,
    >=stealth'
  },
  cov/.style = {
    <->,
    thick,
    >=stealth'
  },
]
  \node[manifest] (y) {$y$};
  \node[manifest] (x) [right = of y] {$x$};
  \node[mean] (1) [above = of y] {1};
  \node[error] (varepsilon) [left = of y] {$\varepsilon$};
  \draw[path] (1) to node[right] {$\alpha$} (y);
  \draw[path] (1) to node[above] {$\mu_x$} (x);
  \draw[path] (x) to node[above] {$\beta$} (y);
  \draw[path] (varepsilon) to node[above] {$1$} (y);
  \draw[cov] (1) to[out=70,in=110,looseness=6] node[above] {$1$} (1);
  \draw[cov] (x) to[out=70,in=110,looseness=5] node[above] {$\sigma_{x}^{2}$} (x);
  \draw[cov] (varepsilon) to[out=70,in=110,looseness=5] node[above] {$\sigma_{\varepsilon}^{2}$} (varepsilon);
\end{tikzpicture}
```

## Symbolic

Let $\left\{ y, x, \varepsilon \right\}$ be the variables of interest.

```{r, echo=FALSE}
u <- matrix(
  data = c(
    paste(expression(alpha)),
    paste(expression(mu[x])),
    0
  )
)
v <- matrix(
  data = c(
    paste(expression(alpha + beta * mu[x])),
    paste(expression(mu[x])),
    0
  )
)
A <- S <- matrix(
  data = 0,
  nrow = dim(u)[1],
  ncol = dim(u)[1]
)
I <- diag(dim(A)[1])
A[1, 2] <- paste(expression(beta))
A[1, 3] <- 1
diag(S) <- c(
  0,
  paste(expression(sigma[x]^2)),
  paste(expression(sigma[varepsilon]^2))
)
filter <- matrix(
  data = c(
    1,
    0,
    0,
    1,
    0,
    0
  ),
  nrow = 2,
  ncol = dim(A)[2]
)
rownames(u) <- rownames(v) <- rownames(A) <- rownames(S) <- colnames(A) <- colnames(S) <- colnames(filter) <- c(
  "y",
  "x",
  "e"
)
colnames(u) <- "u"
rownames(filter) <- c(
  "y",
  "x"
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{A} &=",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{S} &=",
  Ryacas::tex(Ryacas::ysym(S)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{C} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{S} \\left[ \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\right]^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  "\\mathbf{E} \\mathbf{S} \\mathbf{E}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A))),
  Ryacas::tex(Ryacas::ysym(S)),
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A))),
  "^{\\mathsf{T}}",
  "\\\\ &=",
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A)) * Ryacas::ysym(S) * t(solve(Ryacas::ysym(I) - Ryacas::ysym(A)))),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{F} &=",
  Ryacas::tex(Ryacas::ysym(filter)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{M} &=",
  "\\mathbf{F} \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{S} \\left[ \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\right]^{\\mathsf{T}} \\mathbf{F}^{\\mathsf{T}}",
  "\\\\ &=",
  "\\mathbf{F} \\mathbf{E} \\mathbf{S} \\mathbf{E}^{\\mathsf{T}} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  "\\mathbf{F} \\mathbf{C} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(Ryacas::ysym(filter)),
  Ryacas::tex(
    solve(
      Ryacas::ysym(I) - Ryacas::ysym(A)
    ) * Ryacas::ysym(S) * t(
      solve(
        Ryacas::ysym(I) - Ryacas::ysym(A)
      )
    )
  ),
  Ryacas::tex(Ryacas::ysym(filter)),
  "^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(
    Ryacas::ysym(filter) * solve(
      Ryacas::ysym(I) - Ryacas::ysym(A)
    ) * Ryacas::ysym(S) * t(
      solve(Ryacas::ysym(I) - Ryacas::ysym(A))
    ) * t(
      Ryacas::ysym(filter)
    )
  ),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{v} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{u}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]^{\\mathsf{-1}}",
  Ryacas::tex(Ryacas::ysym(u)),
  "\\\\ &=",
  Ryacas::tex(ramR::v_sym(A, u)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{u} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right) \\mathbf{v}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]",
  Ryacas::tex(Ryacas::ysym(v)),
  "\\\\ &=",
  Ryacas::tex(ramR::u_sym(A, v)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{g} &=",
  "\\mathbf{F} \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{u}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]^{-1}",
  Ryacas::tex(Ryacas::ysym(u)),
  "\\\\ &=",
  Ryacas::tex(ramR::g_sym(A, u, filter)),
  "\\end{align*}",
  sep = ""
)
```

### Using the `ramR` Package

```{r}
A
```

```{r}
S
```

```{r}
u
```

```{r}
filter
```

The covariance expectations
can be symbolically derived using the `ramR::C_sym()` function.

```{r}
ramR::C_sym(A, S)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{C} =",
  Ryacas::tex(ramR::C_sym(A, S)),
  "\\end{equation*}",
  sep = ""
)
```

The covariance expectations for the observed variables
can be symbolically derived using the `ramR::M_sym()` function.

```{r}
ramR::M_sym(A, S, filter)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{M} =",
  Ryacas::tex(ramR::M_sym(A, S, filter)),
  "\\end{equation*}",
  sep = ""
)
```

The mean expectations
can be symbolically derived using the `ramR::v_sym()` function.

```{r}
ramR::v_sym(A, u)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{v} =",
  Ryacas::tex(ramR::v_sym(A, u)),
  "\\end{equation*}",
  sep = ""
)
```

The mean expectations for the observed variables
can be symbolically derived using the `ramR::g_sym()` function.

```{r}
ramR::g_sym(A, u, filter)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{g} =",
  Ryacas::tex(ramR::g_sym(A, u, filter)),
  "\\end{equation*}",
  sep = ""
)
```

## Numerical Example

```{r, echo=FALSE}
data_t <- function(n1,
                   n2,
                   mean1,
                   mean2,
                   var1,
                   var2) {
  return(
    rbind(
      cbind(
        y = rnorm(n = n1, mean = mean1, sd = sqrt(var1)),
        x = 0
      ),
      cbind(
        y = rnorm(n = n2, mean = mean2, sd = sqrt(var2)),
        x = 1
      )
    )
  )
}
# Population
n1 <- 500
n2 <- 500
mean1 <- 0
mean2 <- 1
var1 <- 1
var2 <- 1
alpha_pop <- mean1
beta_pop <- mean2 - mean1
# Generate Data
set.seed(42)
data <- data_t(
  n1 = n1,
  n2 = n2,
  mean1 = mean1,
  mean2 = mean2,
  var1 = var1,
  var2 = var2
)
df <- as.data.frame(data)
fit <- lm(y ~ x, data = df)
beta <- coef(fit)
alpha <- beta[1]
beta <- beta[2]
sigma2e <- summary(fit)$sigma^2
Sigma <- var(df)
sigma2x <- Sigma[2, 2]
mu <- colMeans(df)
mu_x <- mu[2]
A <- S <- matrix(
  data = 0,
  nrow = 3,
  ncol = 3
)
A[1, ] <- c(0, beta, 1)
diag(S) <- c(0, sigma2x, sigma2e)
v <- matrix(
  data = c(
    alpha + beta * mu_x,
    mu_x,
    0
  )
)
u <- matrix(
  data = c(
    alpha,
    mu_x,
    0
  )
)
rownames(u) <- rownames(v) <- rownames(A) <- rownames(S) <- colnames(A) <- colnames(S) <- colnames(filter) <- c(
  "y",
  "x",
  "e"
)
colnames(u) <- "u"
rownames(filter) <- c(
  "y",
  "x"
)
```

Let `df` be a random sample with the following parameters

```{r, echo=FALSE}
label <- c(
  "Sample Size",
  "$\\mu$",
  "$\\sigma^2$"
)
group_1 <- c(
  n1,
  mean1,
  var1
)
group_2 <- c(
  n2,
  mean2,
  var2
)
knitr::kable(
  cbind(
    label,
    group_1,
    group_2
  ),
  col.names = c("Parameter", "$x = 0$", "$x = 1$")
)
```

```{r, echo=FALSE}

parameter <- round(
  c(
    alpha_pop,
    beta_pop
  ),
  digits = 2
)
label <- c(
  "$\\alpha$",
  "$\\beta$"
)
description <- c(
  "Mean of $x = 0$.",
  "Mean of $x = 1$ minus $x = 0$."
)
knitr::kable(
  cbind(
    label,
    description,
    parameter
  ),
  col.names = c(
    "Parameter",
    "Description",
    "Estimate"
  )
)
```

```{r}
head(df)
summary(df)
```

```{r, echo=FALSE}
boxplot(
  y ~ x,
  data = df
)
hist(
  df[which(df$x == 0), ][["y"]],
  xlab = expression(y),
  main = expression(paste("Histogram of ", y, " for ", x == 0))
)
hist(
  df[which(df$x == 1), ][["y"]],
  xlab = expression(y),
  main = expression(paste("Histogram of ", y, " for ", x == 1))
)
```

### $t$-test

```{r}
t.test(y ~ x, data = df)
```

### Linear Regression

```{r}
summary(lm(y ~ x, data = df))
```

### Structural Equation Modeling

#### lavaan [@lavaan2012]


```{r}
model <- "
  y ~ x
"
fit <- lavaan::sem(
  model,
  data = df,
  meanstructure = TRUE,
  fixed.x = FALSE
)
lavaan::summary(fit)
```

#### OpenMx [@OpenMx2020]

RAM matrices can be used to specify models in `OpenMx`.
Note, however, that the `u` vector in the RAM notation is
`M` in the `OpenMx` notation.

```{r}
mxData <- OpenMx::mxData(
  observed = df,
  type = "raw"
)
mxA <- OpenMx::mxMatrix(
  type = "Full",
  nrow = 3,
  ncol = 3,
  free = c(
    F, T, F,
    F, F, F,
    F, F, F
  ),
  values = c(
    0, 0.20, 1,
    0, 0, 0,
    0, 0, 0
  ),
  labels = c(
    NA, "beta", NA,
    NA, NA, NA,
    NA, NA, NA
  ),
  byrow = TRUE,
  name = "mxA"
)
mxS <- OpenMx::mxMatrix(
  type = "Symm",
  nrow = 3,
  ncol = 3,
  free = c(
    F, F, F,
    F, T, F,
    F, F, T
  ),
  values = c(
    0, 0, 0,
    0, 0.20, 0,
    0, 0, 0.20
  ),
  labels = c(
    NA, NA, NA,
    NA, "sigma2x", NA,
    NA, NA, "sigma2e"
  ),
  byrow = TRUE,
  name = "mxS"
)
mxM <- OpenMx::mxMatrix(
  type = "Full",
  nrow = 1,
  ncol = 3,
  free = c(
    T, T, F
  ),
  values = c(
    0.20,
    0.20,
    0
  ),
  labels = c(
    "alpha",
    "mux",
    NA
  ),
  byrow = TRUE,
  name = "mxM"
)
mxF <- OpenMx::mxMatrix(
  type = "Full",
  nrow = 2,
  ncol = 3,
  free = FALSE,
  values = c(
    1, 0, 0,
    0, 1, 0
  ),
  byrow = TRUE,
  name = "mxF"
)
expRAM <- OpenMx::mxExpectationRAM(
  A = "mxA",
  S = "mxS",
  F = "mxF",
  M = "mxM",
  dimnames = c(
    "y",
    "x",
    "e"
  )
)
objML <- OpenMx::mxFitFunctionML()
mxMod <- OpenMx::mxModel(
  name = "Model",
  data = mxData,
  matrices = list(
    mxA,
    mxS,
    mxF,
    mxM
  ),
  expectation = expRAM,
  fitfunction = objML
)
fit <- OpenMx::mxRun(mxMod)
summary(fit)
```

### Using the `ramR` Package

```{r}
A
```

```{r}
S
```

```{r}
u
```

```{r}
filter
```

The covariance expectations
can be numerically derived using the `ramR::C_num()` function.

```{r}
ramR::C_num(A, S)
```

The covariance expectations for the observed variables
can be numerically derived using the `ramR::M_num()` function.

```{r}
ramR::M_num(A, S, filter)
```

The mean expectations
can be numerically derived using the `ramR::v_num()` function.

```{r}
ramR::v_num(A, u)
```

The mean expectations for the observed variables
can be numerically derived using the `ramR::v_num()` function.

```{r}
ramR::g_num(A, u, filter)
```

## Equations to RAM

The `ramR` package has a utility function
to convert structural equations to RAM notation.
The Student's $t$-test can be expressed in the following equations

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 LABEL
  e           by        y         1
  y           on        x         beta
  e           with      e         sigma[varepsilon]^2
  x           with      x         sigma[x]^2
  y           on        1         alpha
  x           on        1         mu[x]
"
```

```{tikz, ram-t-02, echo=FALSE, fig.width=4, fig.cap="Student's $t$-test's Structural Equations", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", cache=TRUE}
\usetikzlibrary{er, arrows, positioning}
\begin{tikzpicture}[
  auto,
  node distance = 5mm and 15mm,
  manifest/.style = {
    rectangle,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  latent/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm,
    align = center
  },
  inv/.style = {
    rectangle,
    draw=none,
    fill=none,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  error/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm,
    align = center
  },
  mean/.style = {
    regular polygon,
    regular polygon sides = 3,
    draw,
    thick,
    minimum width = 15mm,
    minimum height = 10mm,
    align = center
  },
  path/.style = {
    ->,
    thick,
    >=stealth'
  },
  cov/.style = {
    <->,
    thick,
    >=stealth'
  },
]
  \node[error] (varepsilon_eq1) {$\varepsilon$};
  \node[manifest] (y_eq1) [right = of varepsilon_eq1] {$y$};
  \node[inv] (label_1) [right = of y_eq1] {1};
  \node[manifest] (y_eq2) [below = of varepsilon_eq1] {$y$};
  \node[manifest] (x_eq2) [below = of y_eq1] {$x$};
  \node[inv] (label_2) [below = of label_1] {$\beta$};
  \node[error] (varepsilon_eq3_1) [below = of y_eq2] {$\varepsilon$};
  \node[error] (varepsilon_eq3_2) [below = of x_eq2] {$\varepsilon$};
  \node[inv] (label_3) [below = of label_2] {$\sigma_{\varepsilon}^{2}$};
  \node[manifest] (x_eq4_1) [below = of varepsilon_eq3_1] {$x$};
  \node[manifest] (x_eq4_2) [below = of varepsilon_eq3_2] {$x$};
  \node[inv] (label_4) [below = of label_3] {$\sigma_{x}^{2}$};
  \node[manifest] (y_eq5) [below = of x_eq4_1] {$y$};
  \node[mean] (1_eq5) [right = of y_eq5] {1};
  \node[inv] (label_5) [below = of label_4] {$\alpha$};
  \node[manifest] (x_eq6) [below = of y_eq5] {$x$};
  \node[mean] (1_eq6) [right = of x_eq6] {1};
  \node[inv] (label_6) [below = of label_5] {$\mu_x$};
  \draw[path] (varepsilon_eq1) to node[above] {by} (y_eq1);
  \draw[path] (x_eq2) to node[above] {on} (y_eq2);
  \draw[cov] (varepsilon_eq3_1) to node[above] {with} (varepsilon_eq3_2);
  \draw[cov] (x_eq4_1) to node[above] {with} (x_eq4_2);
  \draw[path] (1_eq5) to node[above] {on} (y_eq5);
  \draw[path] (1_eq6) to node[above] {on} (x_eq6);
\end{tikzpicture}
```

The error term is treated as a latent variable
and defined with the operation `by`.
It's value is constrained to $1$.
The regression of $y$ on $x$ is defined by operation `on`.
It is labeled as `beta`.
The variance of $x$ and the error variance
are defined using the operation `with`.
These are labeled `sigma[x]^2` and `sigma[varepsilon]^2` respectively.
The intercept and the mean of $x$ are defined using the operation `on 1`.
These are labeled `alpha` and `mu[x]` respectively.

The `ramR::eq2ram` converts the equations to RAM notation.

```{r}
ramR::eq2ram(eq)
```

## Equations to Expectations

The `ramR` package has a utility function
to convert structural equations to expectations
both symbolically and numerically.

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 LABEL
  e           by        y         1
  y           on        x         beta
  e           with      e         sigma[varepsilon]^2
  x           with      x         sigma[x]^2
  y           on        1         alpha
  x           on        1         mu[x]
"
```

```{r}
ramR::eq2exp_sym(eq)
```

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 VALUE
  e           by        y         1.00
  y           on        x         1.00
  e           with      e         1.00
  x           with      x         0.25
  y           on        1         0.00
  x           on        1         0.50
"
```

```{r}
ramR::eq2exp_num(eq)
```
