# One-Way Analysis of Variance {#ram-anova}

In this section,
one-way analysis of variance is presented as a structural equation model
using the RAM notation.
Let $y$ be a continuous dependent variable,
$x$ be a categorical independent variable
with three levels
$\left( x = \{0, 1, 2\} \right)$.
The dependent variable $x$ can be dummy coded as

```{r, echo=FALSE}
x <- c(
  "$x = 0$",
  "$x = 1$",
  "$x = 2$"
)
x1 <- c(
  0,
  1,
  0
)
x2 <- c(
  0,
  0,
  1
)
knitr::kable(
  cbind(
    x, x1, x2
  ),
  col.names = c(
    "$x$",
    "$x_1$",
    "$x_2$"
  ),
  escape = FALSE
)
```

$\varepsilon$ be the stochastic error term
with mean 0 and constant variance of $\sigma_{\varepsilon}^{2}$
across the values of the regressors.
The associations of the variables are given by

\begin{equation*}
  y
  =
  \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \varepsilon
\end{equation*}

\noindent where

- $\beta_0$ is the expected value of $y$ when $x = 0$
- $\beta_1$ is the unit change in $y$ for unit change in $x_1$ while $x_2$ is constant
- $\beta_2$ is the unit change in $y$ for unit change in $x_2$ while $x_1$ is constant
- $\beta_0 + \beta_1$ is the expected value of $y$ when $x = 1$
- $\beta_0 + \beta_2$ is the expected value of $y$ when $x = 2$

```{tikz, ram-aov-01, echo=FALSE, fig.cap="One-Way Analysis of Variance", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", cache=TRUE}
\usetikzlibrary{er, arrows, positioning}
\begin{tikzpicture}[
  auto,
  node distance = 15mm,
  manifest/.style = {
    rectangle,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  latent/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 15mm,
    align = center
  },
  inv/.style = {
    rectangle,
    draw=none,
    fill=none,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  error/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 7mm,
    align = center
  },
  mean/.style={
    regular polygon,
    regular polygon sides = 3,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 7mm,
    minimum height = 7mm
  },
  path/.style = {
    ->,
    thick,
    >=stealth'
  },
  cov/.style = {
    <->,
    thick,
    >=stealth'
  },
]
  \node[manifest] (y) {$y$};
  \node[manifest] (x_1) [above right = of y] {$x_1$};
  \node[manifest] (x_2) [below right = of y] {$x_2$};
  \node[inv] (inv) [right = of y] {};
  \node[mean] (1) [right = of inv] {1};
  \node[error] (varepsilon) [left = of y] {$\varepsilon$};
  \draw[path] (1) to node[above] {$\beta_0$} (y);
  \draw[path] (1) to node[above] {$\mu_{x_1}$} (x_1);
  \draw[path] (1) to node[below] {$\mu_{x_2}$} (x_2);
  \draw[path] (x_1) to node[above] {$\beta_1$} (y);
  \draw[path] (x_2) to node[below] {$\beta_2$} (y);
  \draw[path] (varepsilon) to node[above] {$1$} (y);
  \draw[cov] (1) to[out=70,in=110,looseness=6] node[above] {$1$} (1);
  \draw[cov] (x_1) to[out=70,in=110,looseness=5] node[above] {$\sigma_{a}^{2}$} (x_1);
  \draw[cov] (x_2) to[out=-70,in=-110,looseness=5] node[below] {$\sigma_{b}^{2}$} (x_2);
  \draw[cov] (varepsilon) to[out=70,in=110,looseness=5] node[above] {$\sigma_{\varepsilon}^{2}$} (varepsilon);
\end{tikzpicture}
```

## Symbolic

Let $\left\{ y, x_1, x_2, \varepsilon \right\}$ be the variables of interest.

```{r, echo=FALSE}
u <- matrix(
  data = c(
    paste(expression(beta[0])),
    paste(expression(mu[x1])),
    paste(expression(mu[x2])),
    0
  )
)
v <- matrix(
  data = c(
    paste(expression(beta[0] + beta[1] * mu[x1] + beta[2] * mu[x2])),
    paste(expression(mu[x1])),
    paste(expression(mu[x2])),
    0
  )
)
A <- S <- matrix(
  data = 0,
  nrow = dim(u)[1],
  ncol = dim(u)[1]
)
I <- diag(dim(A)[1])
A[1, ] <- c(0, paste(expression(beta[1])), paste(expression(beta[2])), 1)
diag(S) <- c(
  0,
  paste(expression(sigma[x1]^2)),
  paste(expression(sigma[x2]^2)),
  paste(expression(sigma[varepsilon]^2))
)
filter <- diag(dim(A)[1] - 1)
filter <- cbind(filter, 0)
rownames(u) <- rownames(v) <- rownames(A) <- rownames(S) <- colnames(A) <- colnames(S) <- colnames(filter) <- c(
  "y",
  "x1",
  "x2",
  "e"
)
colnames(u) <- "u"
rownames(filter) <- c(
  "y",
  "x1",
  "x2"
)
```


```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{A} &=",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{S} &=",
  Ryacas::tex(Ryacas::ysym(S)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{C} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{S} \\left[ \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\right]^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  "\\mathbf{E} \\mathbf{S} \\mathbf{E}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A))),
  Ryacas::tex(Ryacas::ysym(S)),
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A))),
  "^{\\mathsf{T}}",
  "\\\\ &=",
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A)) * Ryacas::ysym(S) * t(solve(Ryacas::ysym(I) - Ryacas::ysym(A)))),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{F} &=",
  Ryacas::tex(Ryacas::ysym(filter)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{M} &=",
  "\\mathbf{F} \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{S} \\left[ \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\right]^{\\mathsf{T}} \\mathbf{F}^{\\mathsf{T}}",
  "\\\\ &=",
  "\\mathbf{F} \\mathbf{E} \\mathbf{S} \\mathbf{E}^{\\mathsf{T}} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  "\\mathbf{F} \\mathbf{C} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(Ryacas::ysym(filter)),
  Ryacas::tex(
    solve(
      Ryacas::ysym(I) - Ryacas::ysym(A)
    ) * Ryacas::ysym(S) * t(
      solve(
        Ryacas::ysym(I) - Ryacas::ysym(A)
      )
    )
  ),
  Ryacas::tex(Ryacas::ysym(filter)),
  "^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(
    Ryacas::ysym(filter) * solve(
      Ryacas::ysym(I) - Ryacas::ysym(A)
    ) * Ryacas::ysym(S) * t(
      solve(Ryacas::ysym(I) - Ryacas::ysym(A))
    ) * t(
      Ryacas::ysym(filter)
    )
  ),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{v} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{u}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]^{\\mathsf{-1}}",
  Ryacas::tex(Ryacas::ysym(u)),
  "\\\\ &=",
  Ryacas::tex(ramR::v_sym(A, u)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{u} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right) \\mathbf{v}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]",
  Ryacas::tex(Ryacas::ysym(v)),
  "\\\\ &=",
  Ryacas::tex(Ryacas::simplify(ramR::u_sym(A, v))),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{g} &=",
  "\\mathbf{F} \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{u}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]^{-1}",
  Ryacas::tex(Ryacas::ysym(u)),
  "\\\\ &=",
  Ryacas::tex(ramR::g_sym(A, u, filter)),
  "\\end{align*}",
  sep = ""
)
```

### Using the `ramR` Package

```{r}
A
```

```{r}
S
```

```{r}
u
```

```{r}
filter
```

The covariance expectations
can be symbolically derived using the `ramR::C_sym()` function.

```{r}
ramR::C_sym(A, S)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{C} =",
  Ryacas::tex(ramR::C_sym(A, S)),
  "\\end{equation*}",
  sep = ""
)
```

The covariance expectations for the observed variables
can be symbolically derived using the `ramR::M_sym()` function.

```{r}
ramR::M_sym(A, S, filter)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{M} =",
  Ryacas::tex(ramR::M_sym(A, S, filter)),
  "\\end{equation*}",
  sep = ""
)
```

The mean expectations
can be symbolically derived using the `ramR::v_sym()` function.

```{r}
ramR::v_sym(A, u)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{v} =",
  Ryacas::tex(ramR::v_sym(A, u)),
  "\\end{equation*}",
  sep = ""
)
```

The mean expectations for the observed variables
can be symbolically derived using the `ramR::g_sym()` function.

```{r}
ramR::g_sym(A, u, filter)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{g} =",
  Ryacas::tex(ramR::g_sym(A, u, filter)),
  "\\end{equation*}",
  sep = ""
)
```

## Numerical Example

```{r, echo=FALSE}
data_aov <- function(n,
                     mean,
                     var,
                     dummy = FALSE) {
  if (!all.equal(length(n), length(mean), length(var))) {
    stop(
      "`n`, `mean`, and `var` should have the same length."
    )
  }
  levels <- seq_along(n) - 1
  y <- as.vector(
    mapply(
      FUN = rnorm,
      n = n,
      mean = mean,
      sd = sqrt(var),
      SIMPLIFY = TRUE
    )
  )
  x <- as.vector(
    mapply(
      FUN = rep,
      x = levels,
      times = n,
      SIMPLIFY = TRUE
    )
  )
  if (!dummy) {
    return(
      data.frame(
        y = y,
        x = as.factor(x)
      )
    )
  }
  dummy <- levels[-1]
  dummy_list <- vector(mode = "list", length = length(dummy))
  names(dummy_list) <- paste0("x", dummy)
  for (i in seq_along(dummy)) {
    dummy_list[[i]] <- ifelse(test = x == dummy[i], yes = 1, 0)
  }
  dummy_list <- do.call(
    what = "cbind",
    args = dummy_list
  )
  return(
    cbind(
      y,
      dummy_list
    )
  )
}
# Population
p <- 3
n <- rep(x = 500, times = p)
mean <- c(0, 2, 1)
var <- rep(x = 1, times = p)
beta_0_pop <- mean[1]
beta_1_pop <- mean[2] - mean[1]
beta_2_pop <- mean[3] - mean[1]
# Generate Data
set.seed(42)
data_dummy <- data_aov(n, mean, var, dummy = TRUE)
data <- data_aov(n, mean, var, dummy = FALSE)
df_dummy <- as.data.frame(data_dummy)
df <- as.data.frame(data)
fit <- lm(y ~ x1 + x2, data = df_dummy)
beta <- coef(fit)
beta_0 <- beta[1]
beta_1 <- beta[2]
beta_2 <- beta[3]
sigma2e <- summary(fit)$sigma^2
Sigma <- var(df_dummy)
sigma2x_1 <- Sigma[2, 2]
sigma2x_2 <- Sigma[3, 3]
mu <- colMeans(df_dummy)
mu_x_1 <- mu[2]
mu_x_2 <- mu[3]
A <- S <- matrix(
  data = 0,
  nrow = 4,
  ncol = 4
)
A[1, ] <- c(0, beta_1, beta_2, 1)
diag(S) <- c(0, sigma2x_1, sigma2x_2, sigma2e)
v <- matrix(
  data = c(
    beta_0 + beta_1 * mu_x_1 + beta_2 * mu_x_2,
    mu_x_1,
    mu_x_2,
    0
  ),
  ncol = 1
)
u <- matrix(
  data = c(
    mu_x_1,
    mu_x_1,
    mu_x_2,
    0
  ),
  ncol = 1
)
rownames(u) <- rownames(v) <- rownames(A) <- rownames(S) <- colnames(A) <- colnames(S) <- colnames(filter) <- c(
  "y",
  "x1",
  "x2",
  "e"
)
colnames(u) <- "u"
```

Let `df` be a random sample from a population with the following parameters

```{r, echo=FALSE}
label <- c(
  "Sample Size",
  "$\\mathbb{E} \\left( y \\mid x \\right)$",
  "$\\mathrm{Var} \\left( y \\mid x \\right)$"
)
group_1 <- c(
  n[1],
  mean[1],
  var[1]
)
group_2 <- c(
  n[2],
  mean[2],
  var[2]
)
group_3 <- c(
  n[3],
  mean[3],
  var[3]
)
knitr::kable(
  cbind(
    label,
    group_1,
    group_2,
    group_3
  ),
  col.names = c("Parameter", "$x = 0$", "$x = 1$", "$x = 2$"),
  escape = FALSE
)
```

```{r, echo=FALSE}
parameter <- round(
  c(
    beta_0_pop,
    beta_1_pop,
    beta_2_pop
  ),
  digits = 2
)
label <- c(
  "$\\beta_0$",
  "$\\beta_1$",
  "$\\beta_2$"
)
description <- c(
  "$\\mathbb{E} \\left( y \\mid x = 0 \\right)$",
  "$\\mathbb{E} \\left( y \\mid x = 1 \\right) - \\mathbb{E} \\left( y \\mid x = 0 \\right)$",
  "$\\mathbb{E} \\left( y \\mid x = 2 \\right) - \\mathbb{E} \\left( y \\mid x = 0 \\right)$"
)
knitr::kable(
  cbind(
    label,
    description,
    parameter
  ),
  col.names = c(
    "Parameter",
    "Description",
    "Value"
  ),
  escape = FALSE
)
```

```{r}
head(df)
summary(df)
```

```{r, echo=FALSE}
boxplot(
  y ~ x,
  data = df
)
```

```{r, echo=FALSE}
hist(
  df[which(df$x == 0), ][["y"]],
  xlab = expression(y),
  main = expression(paste("Histogram of ", y, " for ", x == 0))
)
```

```{r, echo=FALSE}
hist(
  df[which(df$x == 1), ][["y"]],
  xlab = expression(y),
  main = expression(paste("Histogram of ", y, " for ", x == 1))
)
```

```{r, echo=FALSE}
hist(
  df[which(df$x == 2), ][["y"]],
  xlab = expression(y),
  main = expression(paste("Histogram of ", y, " for ", x == 2))
)
```

### One-Way Analysis of Variance

Make sure that $x$ is of class `factor` for `lm` and `aov`
to treat it as a categorical variable.

```{r}
str(df)
```

```{r}
summary(aov(y ~ x, data = df))
```

### Linear Regression


```{r}
summary(lm(y ~ x, data = df))
```

### Structural Equation Modeling

We have to dummy code the data set first before fitting the model.
The `model.matrix` function which is used to create a design matrix
can be used to dummy code `x`.
Make sure that `x` is a `factor`.
The first column of the design matrix is a matrix of ones.
Since we do not need this column,
we can replace this column with the values of `y`.
Make sure to name rename the first column
as `lavaan` relies on the column names.

```{r}
df_dummy <- model.matrix(y ~ x, data = df)
df_dummy[, 1] <- df$y
colnames(df_dummy)[1] <- "y"
head(df_dummy)
```

#### lavaan [@lavaan2012]

```{r}
model <- "
  y ~ x1 + x2
"
fit <- lavaan::sem(
  model,
  data = df_dummy,
  meanstructure = TRUE,
  fixed.x = FALSE
)
lavaan::summary(fit)
```

#### OpenMx [@OpenMx2020]

RAM matrices can be used to specify models in `OpenMx`.
Note, however, that the `u` vector in the RAM notation is
`M` in the `OpenMx` notation.

```{r}
mxData <- OpenMx::mxData(
  observed = df_dummy,
  type = "raw"
)
mxA <- OpenMx::mxMatrix(
  type = "Full",
  nrow = 4,
  ncol = 4,
  free = c(
    F, T, T, F,
    F, F, F, F,
    F, F, F, F,
    F, F, F, F
  ),
  values = c(
    0, 0.20, 0.20, 1,
    0, 0, 0, 0,
    0, 0, 0, 0,
    0, 0, 0, 0
  ),
  labels = c(
    NA, "beta1", "beta2", NA,
    NA, NA, NA, NA,
    NA, NA, NA, NA,
    NA, NA, NA, NA
  ),
  byrow = TRUE,
  name = "mxA"
)
mxS <- OpenMx::mxMatrix(
  type = "Symm",
  nrow = 4,
  ncol = 4,
  free = c(
    F, F, F, F,
    F, T, F, F,
    F, F, T, F,
    F, F, F, T
  ),
  values = c(
    0,    0,    0,    0,
    0, 0.20,    0,    0,
    0,    0, 0.20,    0,
    0,    0,    0, 0.20
  ),
  labels = c(
    NA, NA, NA, NA,
    NA, "sigma2x1", NA, NA,
    NA, NA, "sigma2x2", NA,
    NA, NA, NA, "sigma2e"
  ),
  byrow = TRUE,
  name = "mxS"
)
mxM <- OpenMx::mxMatrix(
  type = "Full",
  nrow = 1,
  ncol = 4,
  free = c(
    T, T, T, F
  ),
  values = c(
    0.20,
    0.20,
    0.20,
    0
  ),
  labels = c(
    "beta0",
    "mux1",
    "mux2",
    NA
  ),
  byrow = TRUE,
  name = "mxM"
)
mxF <- OpenMx::mxMatrix(
  type = "Full",
  nrow = 3,
  ncol = 4,
  free = FALSE,
  values = c(
    1, 0, 0, 0,
    0, 1, 0, 0,
    0, 0, 1, 0
  ),
  byrow = TRUE,
  name = "mxF"
)
expRAM <- OpenMx::mxExpectationRAM(
  A = "mxA",
  S = "mxS",
  F = "mxF",
  M = "mxM",
  dimnames = c(
    "y",
    "x1",
    "x2",
    "e"
  )
)
objML <- OpenMx::mxFitFunctionML()
mxMod <- OpenMx::mxModel(
  name = "One Way Analysis of Variance",
  data = mxData,
  matrices = list(
    mxA,
    mxS,
    mxF,
    mxM
  ),
  expectation = expRAM,
  fitfunction = objML
)
fit <- OpenMx::mxRun(mxMod)
summary(fit)
```

### Using the `ramR` Package

```{r}
A
```

```{r}
S
```

```{r}
u
```

```{r}
filter
```

The covariance expectations
can be numerically derived using the `ramR::C_num()` function.

```{r}
ramR::C_num(A, S)
```

The covariance expectations for the observed variables
can be numerically derived using the `ramR::M_num()` function.

```{r}
ramR::M_num(A, S, filter)
```

The mean expectations
can be numerically derived using the `ramR::v_num()` function.

```{r}
ramR::v_num(A, u)
```

The mean expectations for the observed variables
can be numerically derived using the `ramR::v_num()` function.

```{r}
ramR::g_num(A, u, filter)
```

## Equations to RAM

The `ramR` package has a utility function
to convert structural equations to RAM notation.
One-way analysis of variance with three levels
can be expressed in the following equations

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 LABEL
  e           by        y         1
  y           on        x1        beta1
  y           on        x2        beta2
  e           with      e         sigma[varepsilon]^2
  x1          with      x1        sigma[x1]^2
  x2          with      x2        sigma[x2]^2
  y           on        1         beta0
  x1          on        1         mu[x1]
  x2          on        1         mu[x2]
"
```

```{tikz, ram-aov-02, echo=FALSE, fig.cap="One-Way Analysis of Variance's Structural Equations", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", cache=TRUE}
\usetikzlibrary{er, arrows, positioning}
\begin{tikzpicture}[
  auto,
  node distance = 5mm and 15mm,
  manifest/.style = {
    rectangle,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  latent/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm,
    align = center
  },
  inv/.style = {
    rectangle,
    draw=none,
    fill=none,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  error/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm,
    align = center
  },
  mean/.style = {
    regular polygon,
    regular polygon sides = 3,
    draw,
    thick,
    minimum width = 15mm,
    minimum height = 10mm,
    align = center
  },
  path/.style = {
    ->,
    thick,
    >=stealth'
  },
  cov/.style = {
    <->,
    thick,
    >=stealth'
  },
]
  \node[error] (varepsilon_eq1) {$\varepsilon$};
  \node[manifest] (y_eq1) [right = of varepsilon_eq1] {$y$};
  \node[inv] (label_1) [right = of y_eq1] {1};
  \draw[path] (varepsilon_eq1) to node[above] {by} (y_eq1);
  \node[manifest] (y_eq2) [below = of varepsilon_eq1] {$y$};
  \node[manifest] (x1_eq2) [below = of y_eq1] {$x_1$};
  \node[inv] (label_2) [below = of label_1] {$\beta_1$};
  \draw[path] (x1_eq2) to node[above] {on} (y_eq2);
  \node[manifest] (y_eq3) [below = of y_eq2] {$y$};
  \node[manifest] (x2_eq3) [below = of x1_eq2] {$x_2$};
  \node[inv] (label_3) [below = of label_2] {$\beta_2$};
  \draw[path] (x2_eq3) to node[above] {on} (y_eq3);
  \node[error] (varepsilon_eq4_1) [below = of y_eq3] {$\varepsilon$};
  \node[error] (varepsilon_eq4_2) [below = of x2_eq3] {$\varepsilon$};
  \node[inv] (label_4) [below = of label_3] {$\sigma_{\varepsilon}^{2}$};
  \draw[cov] (varepsilon_eq4_1) to node[above] {with} (varepsilon_eq4_2);
  \node[manifest] (x1_eq5_1) [below = of varepsilon_eq4_1] {$x_1$};
  \node[manifest] (x1_eq5_2) [below = of varepsilon_eq4_2] {$x_1$};
  \node[inv] (label_5) [below = of label_4] {$\sigma_{x_1}^{2}$};
  \draw[cov] (x1_eq5_1) to node[above] {with} (x1_eq5_2);
  \node[manifest] (x2_eq6_1) [below = of x1_eq5_1] {$x_2$};
  \node[manifest] (x2_eq6_2) [below = of x1_eq5_2] {$x_2$};
  \node[inv] (label_6) [below = of label_5] {$\sigma_{x_2}^{2}$};
  \draw[cov] (x2_eq6_1) to node[above] {with} (x2_eq6_2);
  \node[manifest] (y_eq7) [below = of x2_eq6_1] {$y$};
  \node[mean] (1_eq7) [right = of y_eq7] {$1$};
  \node[inv] (label_7) [below = of label_6] {$\beta_0$};
  \draw[path] (1_eq7) to node[above] {on} (y_eq7);
  \node[manifest] (x1_eq8) [below = of y_eq7] {$x_1$};
  \node[mean] (1_eq8) [right = of x1_eq8] {$1$};
  \node[inv] (label_8) [below = of label_7] {$\mu_{x_1}$};
  \draw[path] (1_eq8) to node[above] {on} (x1_eq8);
  \node[manifest] (x2_eq9) [below = of x1_eq8] {$x_2$};
  \node[mean] (1_eq9) [right = of x2_eq9] {$1$};
  \node[inv] (label_9) [below = of label_8] {$\mu_{x_2}$};
  \draw[path] (1_eq9) to node[above] {on} (x2_eq9);
\end{tikzpicture}
```

The error term is treated as a latent variable
and defined with the operation `by`.
Its value is constrained to $1$.
The regression of $y$ on $x_1$ and $x_2$ is defined by operation `on`.
The coefficients are labeled as `beta[1]` and `beta[2]` respectively.
The variance of $x_1$, $x_2$ and the error variance
are defined using the operation `with`.
These are labeled `sigma[x1]^2`, `sigma[x2]^2`, and `sigma[varepsilon]^2` respectively.
The intercept and the mean of $x_1$ and $x_2$ are defined using the operation `on 1`.
These are labeled `beta[0]`, `mu[x1]`, and `mu[x2]` respectively.

The `ramR::eq2ram` converts the equations to RAM notation.

```{r}
ramR::eq2ram(eq)
```

## Equations to Expectations

The `ramR` package has a utility function
to convert structural equations to expectations
both symbolically and numerically.

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 LABEL
  e           by        y         1
  y           on        x1        beta1
  y           on        x2        beta2
  e           with      e         sigma[varepsilon]^2
  x1          with      x1        sigma[x1]^2
  x2          with      x2        sigma[x2]^2
  y           on        1         beta0
  x1          on        1         mu[x1]
  x2          on        1         mu[x2]
"
```

```{r}
ramR::eq2exp_sym(eq)
```

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 LABEL
  e           by        y         1
  y           on        x1        2
  y           on        x2        1
  e           with      e         1
  x1          with      x1        0.22222222222
  x2          with      x2        0.22222222222
  y           on        1         0
  x1          on        1         0.33333333333
  x2          on        1         0.33333333333
"
```

```{r}
ramR::eq2exp_num(eq)
```
