# The Simple Mediation Model {#ram-med-simple}

Let $y$, $m$, $x$, $\varepsilon_y$, and $\varepsilon_m$ be random variables
whose associations are given by

\begin{equation}
  y = \beta_0 + \beta_1 x + \beta_2 m + \varepsilon_y
\end{equation}

\begin{equation}
  m = \alpha_0 + \alpha_1 x + \varepsilon_m
\end{equation}

or combined

\begin{equation}
  \begin{split}
    y
    &=
    \beta_0
    + \beta_1 x
    + \beta_2 (\alpha_0 + \alpha_1 x + \varepsilon_m)
    + \varepsilon_y \\
    &=
    \beta_0
    + \beta_1 x
    + \beta_2 \alpha_0
    + \alpha_1 \beta_2 x
    + \beta_2 \varepsilon_m
    + \varepsilon_y
  \end{split}
\end{equation}

where

- $\beta_1$ is the path from $x$ on $y$
- $\beta_2$ is the path from $m$ to $y$
- $\alpha_1$ is the path from $x$ to $m$
- $\varepsilon_y$ and $\varepsilon_m$ are uncorrelated error terms
  with means of zero and constant variances of $\sigma_{\varepsilon_y}^{2}$ and
  $\sigma_{\varepsilon_m}^{2}$ respectively
- $\beta_0$ and $\alpha_0$ are intercepts

```{tikz, ram-med-simple-01, echo=FALSE, fig.cap="The Simple Mediation Model", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", cache=TRUE}
\usetikzlibrary{er, arrows, positioning}
\begin{tikzpicture}[
  auto,
  node distance = 15mm,
  manifest/.style = {
    rectangle,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  latent/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 15mm,
    align = center
  },
  inv/.style = {
    rectangle,
    draw=none,
    fill=none,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  error/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 7mm,
    align = center
  },
  mean/.style={
    regular polygon,
    regular polygon sides = 3,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 7mm,
    minimum height = 7mm
  },
  path/.style = {
    ->,
    thick,
    >=stealth'
  },
  cov/.style = {
    <->,
    thick,
    >=stealth'
  },
]
  \node[manifest] (x) {$x$};
  \node[manifest] (m) [above right = of x] {$m$};
  \node[manifest] (y) [below right = of m] {$y$};
  \node[error] (epsilon_m) [right = of m] {$\varepsilon_m$};
  \node[error] (epsilon_y) [right = of y] {$\varepsilon_y$};
  \draw [path] (x) to node {$\beta_1$} (y);
  \draw [path] (x) to node {$\alpha_1$} (m);
  \draw [path] (m) to node {$\beta_2$} (y);
  \draw [path] (epsilon_m) to node {$1$} (m);
  \draw [path] (epsilon_y) to node {$1$} (y);
  \draw [cov] (x) to[out=170,in=190,looseness=5] node[left] {$\sigma^{2}_{x}$} (x);
  \draw [cov] (epsilon_m) to[out=70,in=110,looseness=5] node[above] {$\sigma^{2}_{\varepsilon_{m}}$} (epsilon_m);
  \draw [cov] (epsilon_y) to[out=70,in=110,looseness=5] node[above] {$\sigma^{2}_{\varepsilon_{y}}$} (epsilon_y);
\end{tikzpicture}
```

```{tikz, ram-med-simple-02, echo=FALSE, fig.cap="The Simple Mediation Model with Mean Structure", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", cache=TRUE}
\usetikzlibrary{er, arrows, positioning}
\begin{tikzpicture}[
  auto,
  node distance = 15mm,
  manifest/.style = {
    rectangle,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  latent/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 15mm,
    align = center
  },
  inv/.style = {
    rectangle,
    draw=none,
    fill=none,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  error/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 7mm,
    align = center
  },
  mean/.style={
    regular polygon,
    regular polygon sides = 3,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 7mm,
    minimum height = 7mm
  },
  path/.style = {
    ->,
    thick,
    >=stealth'
  },
  cov/.style = {
    <->,
    thick,
    >=stealth'
  },
]
  \node[manifest] (y) {$y$};
  \node[manifest] (x) [above right = of y] {$x$};
  \node[manifest] (m) [below right = of y] {$m$};
  \node[inv] (inv) [right = of y] {};
  \node[mean] (1) [right = of inv] {1};
  \node[error] (varepsilon_y) [left = of y] {$\varepsilon_y$};
  \node[error] (varepsilon_m) [below = of m] {$\varepsilon_m$};
  \draw[path] (1) to node[above] {$\beta_0$} (y);
  \draw[path] (1) to node[above] {$\mu_{x}$} (x);
  \draw[path] (1) to node[below] {$\alpha_0$} (m);
  \draw[path] (x) to node[above] {$\beta_1$} (y);
  \draw[path] (x) to node[below left] {$\alpha_1$} (m);
  \draw[path] (m) to node[below] {$\beta_2$} (y);
  \draw[path] (varepsilon_y) to node[above] {$1$} (y);
  \draw[path] (varepsilon_m) to node[right] {$1$} (m);
  \draw[cov] (1) to[out=70,in=110,looseness=6] node[above] {$1$} (1);
  \draw[cov] (x) to[out=70,in=110,looseness=5] node[above] {$\sigma_{x}^{2}$} (x);
  \draw[cov] (varepsilon_y) to[out=70,in=110,looseness=5] node[above] {$\sigma_{\varepsilon_y}^{2}$} (varepsilon_y);
  \draw[cov] (varepsilon_m) to[out=-70,in=-110,looseness=5] node[below] {$\sigma_{\varepsilon_m}^{2}$} (varepsilon_m);
\end{tikzpicture}
```

## Symbolic

Let $\left\{ y, m, x, \varepsilon_y, \varepsilon_m \right\}$ be the variables of interest.

```{r, echo=FALSE}
u <- matrix(
  data = c(
    "beta[0]",
    "alpha[0]",
    "mu[x]",
    0,
    0
  )
)
v <- matrix(
  data = c(
    "beta[0] + beta[1] * alpha[0] + beta[1] * alpha[1] * mu[x] + beta[2] * mu[x]",
    "alpha[0] + alpha[1] * mu[x]",
    "mu[x]",
    0,
    0
  )
)
A <- S <- matrix(
  data = 0,
  nrow = dim(u)[1],
  ncol = dim(u)[1]
)
I <- diag(dim(A)[1])
A[1, ] <- c(0, "beta[2]", "beta[1]", 1, 0)
A[2, ] <- c(0, 0, "alpha[1]", 0, 1)
diag(S) <- c(
  0,
  0,
  "sigma[x]^2",
  "sigma[varepsilon[y]]^2",
  "sigma[varepsilon[m]]^2"
)
filter <- diag(dim(A)[1] - 2)
filter <- cbind(filter, 0, 0)
rownames(u) <- rownames(v) <- rownames(A) <- rownames(S) <- colnames(A) <- colnames(S) <- colnames(filter) <- c(
  "y",
  "m",
  "x",
  "ey",
  "em"
)
colnames(u) <- "u"
rownames(filter) <- c(
  "y",
  "m",
  "x"
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{A} &=",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{S} &=",
  Ryacas::tex(Ryacas::ysym(S)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{C} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{S} \\left[ \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\right]^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  "\\mathbf{E} \\mathbf{S} \\mathbf{E}^{\\mathsf{T}} \\\\",
  "\\\\ &=",
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A))),
  Ryacas::tex(Ryacas::ysym(S)),
  Ryacas::tex(solve(Ryacas::ysym(I) - Ryacas::ysym(A))),
  "^{\\mathsf{T}}",
  "\\\\ &=",
  Ryacas::tex(Ryacas::simplify(solve(Ryacas::ysym(I) - Ryacas::ysym(A)) * Ryacas::ysym(S) * t(solve(Ryacas::ysym(I) - Ryacas::ysym(A))))),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{F} &=",
  Ryacas::tex(Ryacas::ysym(filter)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{M} &=",
  "\\mathbf{F} \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{S} \\left[ \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\right]^{\\mathsf{T}} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "&=",
  "\\mathbf{F} \\mathbf{E} \\mathbf{S} \\mathbf{E}^{\\mathsf{T}} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "&=",
  "\\mathbf{F} \\mathbf{C} \\mathbf{F}^{\\mathsf{T}} \\\\",
  "&=",
  Ryacas::tex(
    Ryacas::simplify(
      Ryacas::ysym(filter) * solve(
        Ryacas::ysym(I) - Ryacas::ysym(A)
      ) * Ryacas::ysym(S) * t(
        solve(Ryacas::ysym(I) - Ryacas::ysym(A))
      ) * t(
        Ryacas::ysym(filter)
      )
    )
  ),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{v} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{u}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]^{\\mathsf{-1}}",
  Ryacas::tex(Ryacas::ysym(u)),
  "\\\\ &=",
  Ryacas::tex(ramR::v(Ryacas::ysym(A), u, simplify = TRUE)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{u} &=",
  "\\left( \\mathbf{I} - \\mathbf{A} \\right) \\mathbf{v}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]",
  Ryacas::tex(Ryacas::ysym(v)),
  "\\\\ &=",
  Ryacas::tex(ramR::u(Ryacas::ysym(A), v, simplify = TRUE)),
  "\\end{align*}",
  sep = ""
)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{align*}",
  "\\mathbf{g} &=",
  "\\mathbf{F} \\left( \\mathbf{I} - \\mathbf{A} \\right)^{-1} \\mathbf{u}",
  "\\\\ &=",
  "\\left[",
  Ryacas::tex(Ryacas::ysym(I)),
  "-",
  Ryacas::tex(Ryacas::ysym(A)),
  "\\right]^{-1}",
  Ryacas::tex(Ryacas::ysym(u)),
  "\\\\ &=",
  Ryacas::tex(ramR::g(Ryacas::ysym(A), u, filter, simplify = TRUE)),
  "\\end{align*}",
  sep = ""
)
```


### Using the `ramR` Package

```{r}
A
```

```{r}
S
```

```{r}
u
```

```{r}
filter
```

The covariance expectations
can be symbolically derived using the `ramR::C()` function
with `A` of class `yac.symbol`.

```{r}
ramR::C(Ryacas::ysym(A), S, simplify = TRUE)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{C} =",
  Ryacas::tex(ramR::C(Ryacas::ysym(A), S, simplify = TRUE)),
  "\\end{equation*}",
  sep = ""
)
```

The covariance expectations for the observed variables
can be symbolically derived using the `ramR::M()` function
with `A` of class `yac.symbol`.

```{r}
ramR::M(Ryacas::ysym(A), S, filter, simplify = TRUE)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{M} =",
  Ryacas::tex(ramR::M(Ryacas::ysym(A), S, filter, simplify = TRUE)),
  "\\end{equation*}",
  sep = ""
)
```

The mean expectations
can be symbolically derived using the `ramR::v()` function
with `A` of class `yac.symbol`.

```{r}
ramR::v(Ryacas::ysym(A), u, simplify = TRUE)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{v} =",
  Ryacas::tex(ramR::v(Ryacas::ysym(A), u, simplify = TRUE)),
  "\\end{equation*}",
  sep = ""
)
```

The mean expectations for the observed variables
can be symbolically derived using the `ramR::g()` function
with `A` of class `yac.symbol`.

```{r}
ramR::g(Ryacas::ysym(A), u, filter, simplify = TRUE)
```

```{r, echo=FALSE, results = "asis"}
cat(
  "\\begin{equation*}",
  "\\mathbf{g} =",
  Ryacas::tex(ramR::g(Ryacas::ysym(A), u, filter, simplify = TRUE)),
  "\\end{equation*}",
  sep = ""
)
```

## Numerical Example

```{r, echo=FALSE}
# Population
sigma2y_pop <- sigma2m_pop <- sigma2x_pop <- 225
mu_y_pop <- mu_m_pop <- mu_x_pop <- 100
alpha_1_pop <- 0.50
beta_1_pop <- 0.00
beta_2_pop <- 0.50
sigma2em_pop <- sigma2m_pop - sigma2x_pop * alpha_1_pop^2
sigma2ey_pop <- sigma2y_pop - sigma2x_pop * beta_2_pop^2 - 2 * sigma2x_pop * beta_2_pop * beta_1_pop * alpha_1_pop - sigma2x_pop * beta_1_pop^2 * alpha_1_pop^2 + beta_1_pop^2 * sigma2em_pop
alpha_0_pop <- mu_m_pop - alpha_1_pop * mu_x_pop
beta_0_pop <- mu_y_pop - beta_1_pop * alpha_0_pop - beta_1_pop * alpha_1_pop * mu_x_pop - beta_2_pop * mu_x_pop
eq <- paste(
  "ey by y 1", "\n",
  "em by m 1", "\n",
  "y on x", beta_1_pop, "\n",
  "y on m", beta_2_pop, "\n",
  "m on x", alpha_1_pop, "\n",
  "ey with ey", sigma2ey_pop, "\n",
  "em with em", sigma2em_pop, "\n",
  "x with x", sigma2x_pop, "\n",
  "y on 1", beta_0_pop, "\n",
  "m on 1", alpha_0_pop, "\n",
  "x on 1", mu_x_pop
)
set.seed(42)
data <- ramR::Eq2Data(eq, n = 1000)
df <- as.data.frame(data)
mod1 <- lm(y ~ x + m, data = df)
mod2 <- lm(m ~ x, data = df)
coef1 <- coef(mod1)
coef2 <- coef(mod2)
beta_0 <- coef1[1]
beta_1 <- coef1[2]
beta_2 <- coef1[3]
alpha_0 <- coef2[1]
alpha_1 <- coef2[2]
sigma2ey <- summary(mod1)$sigma^2
sigma2em <- summary(mod2)$sigma^2
sigma2x <- var(df$x)
mu_x <- mean(df$x)
A <- S <- matrix(
  data = 0,
  nrow = 5,
  ncol = 5
)
A[1, ] <- c(0, beta_2, beta_1, 1, 0)
A[2, ] <- c(0, 0, alpha_1, 0, 1)
diag(S) <- c(
  0,
  0,
  sigma2x,
  sigma2ey,
  sigma2em
)
u <- matrix(
  data = c(
    beta_0,
    alpha_0,
    mu_x,
    0,
    0
  ),
  ncol = 1
)
rownames(u) <- rownames(v) <- rownames(A) <- rownames(S) <- colnames(A) <- colnames(S) <- c(
  "y",
  "m",
  "x",
  "ey",
  "em"
)
colnames(u) <- "u"
```

Let `df` be a random sample from a population with the following parameters

```{r, echo = FALSE}
Parameter <- c(
  "$\\beta_1$",
  "$\\beta_2$",
  "$\\alpha_1$",
  "$\\sigma_{\\varepsilon_y}^{2}$",
  "$\\sigma_{\\varepsilon_m}^{2}$",
  "$\\sigma_{x}^{2}$",
  "$\\beta_0$",
  "$\\alpha_0$",
  "$\\mu_x$"
)
Value <- c(
  beta_1_pop,
  beta_2_pop,
  alpha_1_pop,
  sigma2ey_pop,
  sigma2em_pop,
  sigma2x_pop,
  beta_0_pop,
  alpha_0_pop,
  mu_x_pop
)
knitr::kable(
  cbind(
    Parameter,
    Value
  ),
  escape = FALSE
)
```

```{r}
head(df)
summary(df)
```

```{r, echo=FALSE}
pairs(df)
```

### Linear Regression

```{r}
summary(lm(y ~ x + m, data = df))
summary(lm(m ~ x, data = df))
```

### Structural Equation Modeling

#### lavaan [@lavaan2012]

```{r}
# Covariance Structure
model <- "
  y ~ beta1 * x + beta2 * m
  m ~ alpha1 * x
  indirect := alpha1 * beta2
"
# With Mean Structure
model <- "
  y ~ beta1 * x + beta2 * m
  m ~ alpha1 * x
  y ~~ sigma2ey * y
  m ~~ sigma2em * m
  x ~~ sigma2x * x
  y ~ beta0 * 1
  m ~ alpha0 * 1
  x ~ mux * 1
  indirect := alpha1 * beta2
"
fit <- lavaan::sem(
  model,
  data = df,
  meanstructure = TRUE,
  fixed.x = FALSE
)
lavaan::summary(fit)
```

### Using the `ramR` Package

```{r}
A
```

```{r}
S
```

```{r}
u
```

```{r}
filter
```

The covariance expectations
can be numerically derived using the `ramR::C()` function.

```{r}
ramR::C(A, S)
```

The covariance expectations for the observed variables
can be numerically derived using the `ramR::M()` function.

```{r}
ramR::M(A, S, filter)
```

The mean expectations
can be numerically derived using the `ramR::v()` function.

```{r}
ramR::v(A, u)
```

The mean expectations for the observed variables
can be numerically derived using the `ramR::v()` function.

```{r}
ramR::g(A, u, filter)
```

## Equations to RAM

The `ramR` package has a utility function
to convert structural equations to RAM notation.
The simple mediation model
can be expressed in the following equations

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 LABEL
  ey          by        y         1
  em          by        m         1
  y           on        x         beta[1]
  y           on        m         beta[2]
  m           on        x         alpha[1]
  ey          with      ey        sigma[varepsilon[y]]^2
  em          with      em        sigma[varepsilon[m]]^2
  x           with      x         sigma[x]^2
  y           on        1         beta[0]
  m           on        1         alpha[0]
  x           on        1         mu[x]
"
```

The `ramR::Eq2RAM` converts the equations to RAM notation.

```{r}
ramR::Eq2RAM(eq)
```

## Equations to Expectations

The `ramR` package has a utility function
to convert structural equations to expectations
both symbolically and numerically.

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 LABEL
  ey          by        y         1
  em          by        m         1
  y           on        x         beta[1]
  y           on        m         beta[2]
  m           on        x         alpha[1]
  ey          with      ey        sigma[varepsilon[y]]^2
  em          with      em        sigma[varepsilon[m]]^2
  x           with      x         sigma[x]^2
  y           on        1         beta[0]
  m           on        1         alpha[0]
  x           on        1         mu[x]
"
```

```{r}
ramR::Eq2Expectations(eq)
```

```{r}
eq <- "
  # VARIABLE1 OPERATION VARIABLE2 LABEL
  ey          by        y         1
  em          by        m         1
  y           on        x         0.00
  y           on        m         0.50
  m           on        x         0.50
  ey          with      ey        168.75
  em          with      em        168.75
  x           with      x         225
  y           on        1         50
  m           on        1         50
  x           on        1         100
"
```

```{r}
ramR::Eq2Expectations(eq)
```
